---
title: "TASK (dataset: caerphilly)"
author: "Amanda Wittendorf"
date: "`r Sys.Date()`"
output: html_document
---


```{r Pre-setting_plots, message=FALSE, warning=FALSE, include=FALSE}
# PRE-SETTING FOR THEMES IN PLOTS 
library(ggplot2)

standard_tema <- theme_minimal(base_size = 11, base_family = "sans") + 
  theme(
    plot.title = element_text(face = "bold", 
                              size = rel(1.2), 
                              color = "#333333"), 
    plot.subtitle = element_text(face = "italic", 
                                 size = rel(0.95), 
                                 color = "#555555"), 
    axis.title = element_text(size = rel(0.9)), 
    axis.text = element_text(size = rel(0.85)), 
    legend.position = "bottom"
  )

# Set as global standard-theme
theme_set(standard_tema)
```

## 1. DESKRIPTION ON THE DATASET

The dataset is from the pack `statistisk` and contains measurements Myocardial infarction (MI). More specifically: 

- $\text{id}$:        Individuals ID-number 

- $\text{birthdate}$: Date of birth (Eksempel: 1910-04-10)

- $\text{examdate}$:  Date of first examination (Eksempel: 1980-02-21)

- $\text{dthdate}$:   Date of death (Eksempel: NA)

- $\text{midate}$:    Date of first MI (Eksempel: 1996-02-09)

- $\text{emdate}$:    Date of emigration (Eksempel: NA)

- $\text{eosdate}$:   End of study (Eksempel: 1999-02-14)

- $\text{socclass}$:  Social class based upon The Registrar General's Social Class Scale (=RGSC) in Britian  (I / II / IIINM / IIIM / IV / V)
  - I: Professional occupations (e.g. doctors, layers, etc.)
  - II: Managerial and technical / Intermediate (e.g. nurses, pilots, etc.)
  - IIINM: Skilled non-manual occupations (e.g. receptionist, clerical staff etc.)
  - IIIM: Skilled manual occupations (e.g. bus drivers, hairdresser, etc.)
  - IV: 	Partly skilled occupations (e.g. bartender, janitor, etc.)
  - V: 	Unskilled occupations (e.g. cleaners, piccolo, etc.)
  
- $\text{diabetes}$:  Diabetes status (No/uncertain / Yes)

- $\text{smoking}$:   Smoking status 
                      (<15 per day / Never smoked / Ex>5 years / Ex 1-4 years / >15 per day)  

- $\text{fibrin}$:    Fibrinogen at baseline

- $\text{totchol}$:   Total cholesterol at baseline

- $\text{hdlchol}$:   HDL cholesterol at baseline

- $\text{bpsyst}$:    Systolic blood pressure

- $\text{bpdias}$:    Diastolic blood pressure

- $\text{bmi}$:       Body mass index (kg/m2)

- $\text{logtrig}$:   Log (base 10) triglycerides


```{r dataset, message=FALSE, include=FALSE}
library(statistik)
library(survival)

# Dataset
caer <- Caerphilly |> 
  mutate(socclass = factor(socclass),
         diabetes = factor(diabetes),
         smoking = factor(smoking)
  )
```


## 2. DATA MANAGEMENT 

### 2.1. Definition of Follow-up and Endpoints

To perform an analysis on survival data following variables need to be constructed:

1. **The date of study exit** due to 
  (a) MI, 
  (b) death, 
  (c) emigration, or 
  (d) end-of-study

2. **Status variable** indicating whether or not the individual experienced an MI prior to exit. 

3. **Time variable** describing how many years an individual remains in the study before exiting. 

Death, emigration, and end-of-study are all going to be treated as censoring events.  

```{r new_variables, echo=FALSE, message=FALSE, warning=FALSE}
library(knitr)
library(kableExtra) 

caer <- caer |> 
  # Item 1
  mutate(date_out = pmin(dthdate, emdate, eosdate, midate, 
                         na.rm = TRUE),
  # Item 2 
         status = case_when(
           date_out == midate ~ "MI",
           .default = "Censored"),    # <- Censoring events
         status = factor(status)
         ) |> 
  
  # Item 3
  mutate(time_study = difftime(date_out, examdate, units = "days"),
         time_study = as.numeric(time_study) / 365.25)

# Print 
select(caer,  
       id, status, examdate, midate, dthdate, emdate, eosdate, date_out, time_study) |>
  as.tibble() |> 
  head(8) |> 
  rename("ID" = id,
         "MI (midate)" = midate,
         "Death (dthdate)" = dthdate,
         "Emigration (emdate)" = emdate,
         "End-of-study (eosdate)" = eosdate,
         "First examination (examdate)" = examdate,
         "Event (date_out)" = date_out,
         "Time in years (time_study)" = time_study,
         "Status variabel (status)" = status) |> 
  kable(format = "html", align = "lr", , digits = 2,
        caption = "Table 2.1: Key timeline events and follow-up dates (n = 8)") |> 
  kable_classic(full_width = FALSE, html_font = "Cambria")
```


### 2.2. Definition of Time-at-Risk and Age at inclusion (years)

Time-at-Risk is calculated for each person and is defined as how much risk time they contribute, and whether they experienced the MI event using `pyears` while Age at inclusion is being recoded to be a numeric variable.

```{r time_at_risk, echo=FALSE, message=FALSE, warning=FALSE}
# Time-at-Risk
py <- pyears(Surv(time_study, status == "MI") ~ id,
             scale = 1,
             data.frame = TRUE,
             data = caer)

caer <- py$data |>
  select(-n) |>
  right_join(caer, by = "id") |>
  as_tibble()

# Age at inclusion
caer <- caer |> 
  mutate(age_exam = difftime(examdate, birthdate, units = "days"),
                        age_exam = as.numeric(age_exam) / 365.25)

# Print
select(caer, 
       id, pyears, age_exam, event) |> 
  as.tibble() |> 
  head(8) |> 
  rename("ID" = id,
         "Person-Years (pyears)" = pyears,
         "Age at inclusion (age_exam)" = age_exam,
         "Event of MI (event)" = event
         ) |> 
  kable(format = "html", align = "lr", , digits = 2,
        caption = "Table 2.2: Sample overview of individual follow-up time, event status and age at inclusion (n = 8)") |> 
  kable_classic(full_width = FALSE, html_font = "Cambria") |> 
  footnote(general = "MI: Myocardial infarction. Event status is coded as 1 for individuals who experienced \nan MI during follow-up, and 0 for censored observations.\n
  N = 1786
  Total events = 348
  Total person-years = 26161.2"
  )
```
```{r person_years, message=FALSE, warning=FALSE, include=FALSE}
py
```

**INTERPRETATION**

Table 2.2 shows both the individual as well as the total amount of events, person-years, and age at inclusion (years). 

**Events:**
The total number of individuals who experienced the primary outcome, myocardial infarction (MI), during the follow-up period. The remaining 1,438 individuals did not experience an MI before the study ended or before they dropped out (censoring).

**Person-years:**
The cumulative amount of time all participants contributed while being active in the study and at risk of experiencing an MI. It is equivalent to following a single person for `r round(sum(py$pyears), 1)` years, or following the`r py$n` participants for an average of approximately `r round(sum(py$pyears) / py$n, 1)` years each (`r round(sum(py$pyears), 1)` person-years divided with `r py$n` individuals)

**Age at inclusion:**
Represents the precise age of the participants in years at baseline. 


### 2.3. Definition of Smoker 

The new binary variable $\text{smoker}$ is made from the variable $\text{smoking}$ by collapsing the five initial categories into two: 

- *Current smoker:* <15 per day / >15 per day

- *Non-smoker:* Never smoked / Ex>5 years / Ex 1-4 years

```{r collapse_smoker,echo=FALSE}
caer <- caer |> 
  mutate(smoker = case_when(
    smoking == "<15 per day" ~ "Current smoker",
    smoking == ">15 per day" ~ "Current smoker",
    .default = "Non-smoker"),
    smoker = factor(smoker))

# Print
xtabs(~smoking + smoker, 
      addNA = TRUE, data = caer) |> 
  as.matrix() |> 
  kable(format = "html", align = "rcc",
        caption = "Table 2.3: Cross-tabulation of the original and recoded smoking variable") |> 
  kable_classic(full_width = FALSE, html_font = "Cambria") |> 
    add_header_above(c("Smoking (original)" = 1, 
                       "Smoker (recoded)" = 2))
```

Table 2.3 shows, that the new variable $\text{smoker}$ is correctly collapsed from the original variable $\text{smoking}$. 


### 2.4. Definition of BMI-group 

The baseline dataset contains body mass index (BMI) as a continuous variable. To improve clinical interpretability, this variable was recoded into a categorical variable based on the standardized thresholds defined by the *World Health Organization (WHO)*:

1. **Underweight / Low normal weight (<18.5):** Individuals with a low body weight.

2. **Normal weight (18.5 – 24.9):** The ideal or reference weight range for adults.

3. **Overweight (25.0 – 29.9):** Moderately elevated body weight (pre-obesity).

4. **Obesity (>30.0):** Significantly elevated body weight (adipositas).

For each of these newly established categories, the total sample size (\(n\)) along with the minimum and maximum observed values were calculated to validate the data distribution

```{r bmi_group, echo=FALSE}
caer <- caer |> 
  mutate(bmi_group = cut(bmi, c(0, 18.5, 25, 30, 100), 
                         labels = c("<18.5", "18.5 - 24.9", "25 - 29.9", ">30"), 
                         right = FALSE),
         bmi_group = factor(bmi_group),
         bmi_group = fct_relevel(bmi_group, "<18.5"))

# Print
caer |> 
  summarise(
    "Amount (n)" = n(),
    "Minimum BMI" = min(bmi, na.rm = TRUE),
    "Maximum BMI" = max(bmi, na.rm = TRUE),
    .by = bmi_group
  ) |> 
  arrange(bmi_group) |>
  rename("BMI group" = bmi_group) |> 
  kable(format = "html", align = "lrrr", digits = 2,
        caption = "Table 2.4: Distribution and validation of categorised BMI groups") |> 
  kable_classic(full_width = FALSE, html_font = "Cambria") |> 
  footnote(general = "BMI is measured in kg/m²")
```



## 3. POISSON REGRESSION MODEL 

A poisson regression model estimates the rate ratio (RR). 


### 3.1 THE EFFECT OF SMOKING STATUS ON THE MI-RATE, ADJUSTED FOR BMI-GROUP

To assess the effect of smoking on the MI-rate the reference group for the variable $\text{smoking}$ needs to be recoded so the reference is non-smokers.

```{r recode_smoking, include=FALSE, warning=FALSE}
caer <- caer |> 
  mutate(smoker = fct_relevel(smoker, "Non-smoker"),
         bmi_group = fct_relevel(bmi_group, "18.5 - 24.9"))
```

```{r smoke_bmi_group, message=FALSE, warning=FALSE, include=FALSE}
table_1 <- list(
  "Poisson" = glm(event ~ offset(log(pyears)) + smoker + bmi_group, family = poisson, data = caer)|> parameters(exponentiate = TRUE) |> as.data.frame()
  ) |> 
  bind_rows(.id = "Table 3.1: Poisson regression model") |> 
  select("Table 3.1: Poisson regression model" = Parameter, 
         Rate = Coefficient, CI_low, CI_high, p) |> 
  mutate(
    across(c(Rate, CI_low, CI_high), ~ round(.x, 2)),
    p = case_when(
      p < 0.001 ~ "< 0.001",
      p < 0.05  ~ "< 0.05",
      TRUE      ~ as.character(round(p, 3))
    )
  )

table_1
```

```{r table_poisson, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(stargazer)

## Models
model   <- glm(event ~ offset(log(pyears)) + smoker + bmi_group, family = poisson, data = caer)

## Transforming from log-odds to odds
RR <- exp(coef(model))
CI <- exp(confint(model))
p  <- summary(model)$coefficients[, 4]

## Table
stargazer(model, 
          type = "html",    
          align = FALSE,
          coef = list(RR), 
          title = "Table 3.1: Poisson regression model",
          ci = TRUE,                      
          ci.custom = list(CI),    
          p = list(p),           
          dep.var.labels = "Event of MI (event)",
          covariate.labels = c("Current Smoker", 
                               "Underweight (<18.5)", "Overweight (25 - 29.9)", "Obesity (>30)",
                               "Intercept"),
          digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001), 
          keep.stat = c("n"),
          notes.align = "l",
          notes.append = TRUE,
          notes = c("Reference group: A non-smoker measured to be in the BMI-group 18.5 - 24.9 (normal weight).")
)
```
**INTERPRETATION**

**Intercept:**
The significant rate for a individual in the reference group per 1,000 person-years; `r table_1$Rate[1]` (95% CI: `r table_1$CI_low[1]`; `r table_1$CI_high[1]`, p `r table_1$p[1]`)

**Smoker:**
The rate for smokers are `r table_1$Rate[2]` (95% CI: `r table_1$CI_low[2]`; `r table_1$CI_high[2]`, p `r table_1$p[2]`) times significant higher to develop MI compared to non-smokers when adjusting for BMI-group. 

**Underweight (<18.5):**
The MI-rate is tending to be `r table_1$Rate[3]` (95% CI: `r table_1$CI_low[3]`; `r table_1$CI_high[3]`, p `r table_1$p[3]`) times lower for non-smoking individuals how is classified as being underweight compared to non-smoking individuals classified as being normal weight. This is not statistically significant. 

**Overweight (25 - 29.9):**
The MI-rate is `r table_1$Rate[4]` (95% CI: `r table_1$CI_low[4]`; `r table_1$CI_high[4]`, p `r table_1$p[4]`) times higher for non-smoking individuals how is classified as being overweight compared to non-smoking individuals classified as being normal weight. This is statistically significant. 

**Obesity (>30):**
Same as above...



#### 3.1.1 EFFECT MODIFICATION










